Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2019-07-02 markdown/felony_disenfranchisement_nys.Rmd
write

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
biblio-style: apalike
link-citations: yes
header-includes:
    - \usepackage{setspace}\doublespacing
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\tableofcontents
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

>>>>

---
title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City"
author: "Kevin Morris"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
bibliography: "../research/citations.bib"
link-citations: yes
header-includes:
    - \usepackage{setspace}\doublespacing
---
<!-- --- -->
<!-- title: "Neighborhoods and Felony Disenfranchisement: The Case of New York City" -->
<!-- author: "Kevin Morris" -->
<!-- date: "`r format(Sys.time(), '%B %d, %Y')`" -->
<!-- output: word_document -->
<!-- bibliography: "../research/citations.bib" -->
<!-- link-citations: yes -->
<!-- always_allow_html: yes -->
<!-- --- -->
\pagenumbering{gobble}
\tableofcontents
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
# Introduction {-}

### 2019-07-02 markdown/part2_standalone.Rmd
write

In this paper, I examine the effect of Executive Order 181 in New York State. Signed in April of 2018, the Executive Order restored voting rights to many individuals on parole. Prior to Executive Order 181, individuals were disenfranchised while incarcerated or on parole. Most individuals who were on parole on Election Day in 2018 were allowed to vote, a major step forward for democracy in the Empire State. In this paper, however, I focus on individuals who were discharged from parole prior to Election Day. These individuals, therefore, would not have been directly disenfranchised by felony disenfranchisement policies even in the absense of the Executive Order. I hypothesize that Executive Order 181 increased turnout among this population, even though their eligibility was not directly impacted by the policy change.

@@

Secondly, Executive Order 181 is expected to dispel confusion about eligibilty. Since @Downs1957, political scientists have argued that an individual's propensity to vote is decided at least in part by comparing the expected costs of casting a ballot with the expected benefits of doing so. The costs of voting while disenfranchised are exceedingly high, as high-profile cases in states such as Texas have made clear in recent years [@Flynn2018]. If an individual assessment of the cost of voting is determined in part by the product of their uncertainty about their eligibility and the cost of casting a ballot while ineligible, formerly disenfranchised individals are unlikely to participate unless they are fully certain of their eligibility. As @Drucker2005 and others have detailed, however, many formerly incarcerated individuals are misinformed about their eligibility to cast a ballot. By dispelling uncertainty around eligibility, I hypothesize that Executive Order 181 increased turnout among the formerly disenfranchised.

@@

On April 18^th^, 2018, Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. While the Executive Order is obviously beneficial for individuals who are on parole on Election Day (and therefore would have been disenfranchised but for the policy change), the Executive Order may have benefits even for the individuals whose eligibility was not directly impacted. Under the Executive Order, parole officers were required to inform their parolees of their new status under the law. In the analysis that follows, I look only at individuals who were discharged before October 10^th^, 2018 (the registration deadline in New York State). Although Executive Order 181 did not change the eligibility of these individuals to cast a ballot in November, there is reason to believe that it might still have impacted their turnout propensity.

In order to estimate the impact of Executive Order 181 on turnout, I begin by exploring whether the turnout rate among individuals whose rights were restored prior to discharge from parole systematically differed from those whose rights were not restored. This analysis uses an individual-level logistic regression that estimates individuals' turnout rates and controls for various individual-level characteristics.

@@

## Voter File Data {-}
Most states in the United States are required to maintain files with information on all registered voters. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first, middle, and last name; date of birth; home address; vote history; and other information. The New York State Voter File also includes information on voters who were previously registered but have since been purged, either because they moved, died, or were incarcerated for a felony offense. I use a snapshot of the registered voter file from March 3^rd^, 2019.

## Matching {-}
Turnout in the 2018 midterm election is estimated by matching the parole records with the registered voter file. I match individuals in each dataset using first name, middle name, last name, and date of birth. Of course, matching using administrative data is not a perfect science. Matching on first and last names and dates of birth can result in false positive matches, especially in large states like New York. Following @Meredith2013, I discuss the potential impact of these errors in [Appendix A]. As Appendix A makes clear, however, there is little reason to believe that false matches systematically impact the results of the analyses that follow.

@@

@@

```{r to-16-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-16}Turnout in 2016 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to16_chart.rds")
print(j)
```
Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January, 2012 through April, 2018, and extended through October, 2018.
```{r to-18-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart.rds")
print(j)
```
Figure \ref{fig:to-16} does not indicate that individuals who were discharged from parole shortly before the 2016 presidential election were more likely to cast a ballot than individuals discharged earlier in the year. Figure \ref{fig:to-18}, on the other hand, indicates that New Yorkers discharged from parole in the months leading up to the 2018 election --- many of whom had their rights restored while they were still on parole --- were more likely to participate than those discharged earlier in the year. However, Figures \ref{fig:to-16} and \ref{fig:to-18} are noisy. In the section that follows, I develop a logit model to explore whether, after controlling for the available characteristics, individuals whose rights were restored prior to being discharged were more likely to cast a ballot in the 2018 midterm elections.

@@

Figure \ref{fig:to-18} displays the share of individuals who were discharged from parole in each month and cast a ballot in the 2018 elections. It does not control for any individual level characteristics. In Table \ref{tab:to-18-logit}, I present the results of an individual-level logistic regression exploring turnout in 2018 among individuals whose rights were restored. It includes all individuals discharged from parole between January 1^st^, 2012, through October 10^th^, 2018 (the registration deadline in New York State).

```{r logit-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/individual_turnout_18.rdata")

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Rights Restored)", "Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:to-18-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```

Model 1 in Table \ref{tab:to-18-logit} formalizes the trend presented in Figure \ref{fig:to-18} by controlling only for whether an individual had his rights restored before being discharged from parole, and the number of days between the individual's discharge date and November 6^th^, 2018. Model 2 also controls for individual-level characteristics: sex, age on November 6^th^, 2018, county of incarceration, and race. Model 3 adds sentence-specific information to Model 2: the number of counts in the individual's most recent sentence, the amount of time they spent on parole, and the class(es) of felony for which they were convicted. In each successive model, the AIC decreases substantially, indicating that these controls are warranted. Table \ref{tab:to-18-logit} makes clear that formerly incarcerated men were far less likely to vote than formerly incarcerated women; that older formerly incarcerated individuals were more likely to cast a ballot; and individuals who spent longer on parole were more likely to participate in the midterm election.

Each model also indicates that individuals whose rights were restored prior to discharge were more likely to cast a ballot than those whose rights were not restored. Exponentiating the coefficients on D(Rights Restored) indicates that individuals whose rights were restored due to Executive Order 181 were between `r round((exp(model1[["coefficients"]][["restoredTRUE"]]) - 1) * 100, digits = 1)` and `r round((exp(model3[["coefficients"]][["restoredTRUE"]]) - 1) * 100, digits = 1)` percent more likely to vote than those whose rights were not.

Table \ref{tab:to-18-logit} provides some indication that Executive Order 181 was successful, but does not establish causality. There is reason to believe that individuals who had their rights restored were more likely to vote even had they not been "treated" by EO 181. In the section that follows, I employ an instrumental variables approach to control for this possibility.

## Instrumenting Rights Restoration {-}

In the previous section, I explored whether individuals who had their rights formally restored prior to parole discharge were more likely to vote in the 2018 midterm election. There is reason to believe that whether an individual received the treatment (had their rights restored) is correlated with their propensity to vote. For instance, parolees who were arrested while on parole did not have their rights restored; a recent arrest is also likely to impact turnout. Similarly, noncitizens did not have voting rights restored / granted; such citizenship status is also of course correlated with propensity to vote. Because the treatment was not randomly assigned, direct comparison of turnout rates among individuals who did and did not have their rights restored is not probitive to the causal effect of Executive Order 181. This calls for an instrumental variables approach.

The timing of Executive Order 181 serves as a useful instrument for formal rights restoration. Although Figure \ref{fig:to-18} shows that there is a relationship between how long an individual has been off parole and their propensity to vote over the long run, this relationship is nonsignficant in the short run. Figure \ref{fig:to-18-recent} shows that, for individuals discharged from parole in 2017 and 2018, there does not appear to be a relationship between time off parole and turnout rates.

```{r to-18-recent-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18-recent}Turnout in 2018 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart_recent.rds")
print(j)
```

Formalizing this chart into an individual-level logit model demonstrates that time since discharge is not correlated with turnout in the short run. The models in Table \ref{tab:to-18-logit-short} include individuals last discharged from parole between January 1^st^, 2017, and April 17^th^, 2018.

@@

The inclusion of time controls in Table \ref{tab:to-18-logit-short} increases the AIC. A Chi-squared test confirms that the model is not improved when controls for time are included.^[[Appendix C] provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.] Over the short term, the length of time an individual has been off parole is not associated with his propensity to vote.

Because the date of parole discharge is not correlated with turnout, we can conceptualize the timing of Executive Order 181 as a randomly assigned treatment. An individual who was released after the Executive Order went into effect is part of the intend-to-treat group, and individuals who actually had their rights restored "comply" with the treatment. As with many true randomized control trials, compliance (restoration) is correlated with the outcome of interest (turnout) and therefore must be controlled for.

Governor Cuomo signed Executive Order 181 in April of 2018, but an examination of the individuals whose rights were ultimately restored indicates that the program did not go into full effect until later in May. As Figure \ref{fig:restoration-chart} demonstrates, individuals who finished their parole supervision in early May did not have their voting rights restored before they finished their term of supervision. It is not until May 21^st^ that the majority of parolees had their voting rights restored prior to their discharge.
```{r restoration-chart-chunk, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:restoration-chart}Share of Discharged Parolees Whose Voting Rights Were Restored Prior to Discharge", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/restoration_plot.rds")
print(j)
```
There is no evidence that date of discharged influences propensity to vote through channels other than formal rights restoration in the short run. A dummy variable indicating whether an individual was discharged on or after May 21^st^ is therefore analogous to an intent-to-treat, with a dummy indicating whether one's rights were restored measuring imperfect compliance. Table \ref{tab:iv-models} shows the second-stage estimation results, estimated using an OLS sepcification.^[Although the dependent variable remains dichotomous (as do both the independent variable of interest and the instrument), @Angrist2008 recommends using OLS in instrumental variables analyses even under such conditions.]

@@

Table \ref{tab:iv-models} indicates that using discharge date of before or after May 21^st^ as an instrument for rights restoration is warranted. The p-values of the Wu-Hausman tests hover around 0.05, while the Weak Instruments test is highly significant in each model.

Model 3 in Table \ref{tab:iv-models} indicates that having one's rights restored caused an individual's propensity to vote to increase by `r round(summary(iv3)[["coefficients"]][2,1] * 100, digits = 2)` percentage points. Approximately `r round(readRDS("./temp/to_restored.rds") * 100, digits = 2)` percent of individuals who had their rights restored cast a ballot in 2018, indicating that Executive Order 181 increased the propensity to vote by around `r round(summary(iv3)[["coefficients"]][2,1] / (readRDS("./temp/to_restored.rds") - summary(iv3)[["coefficients"]][2,1]) * 100, digits = 1)` percent. This estimate is much lower than the estimates reported in Table \ref{tab:to-18-logit}, but that is not surprising; as discussed above, individuals whose rights were restored were more likely to vote whether or not they received restoration before discharge. The `r round(summary(iv3)[["coefficients"]][2,1] * 100, digits = 2)` percentage point increase reported in Table \ref{tab:iv-models} accounts only for the extent to which rights restoration prior to discharge from parole caused higher turnout.

## Discussion {-}
As discussed above, the effect of incarceration on individuals' propensity to vote is an open question for the literature. What is known, however, is that individuals who have been to prison were likely to vote at very low rates prior to incarceration, and continue to vote at very low rates after incarceration. Whether incarceration reduces political participation or not, increasing the propensity to vote of individuals who have been to prison is a laudable goal. Prisoners overwhelmingly come from marginalized communities with much to gain from policy. Executive Order 181 appears to have substantially boosted turnout among formerly incarcerated individuals through two mechanisms. The first is obvious: parolees who would not have been eligible to vote absent the Executive Order were allowed to cast ballots in the midterm elections. The second mechanism is less obvious. Individuals who finished parole between May 21^st^, 2018, and  October 10^th^, 2018, would have been eligible to participate in the election even if the governor had not signed the Executive Order. The EO 181, however, appears to have boosted turnout even among individuals whose eligibility to vote was not directly impacted by the rules change. The increase in turnout was substantial: Executive Order 181 increased turnout among individuals who had their rights restored by more than 40 percent.

@@

>>>>

In this paper, I examine the effect of Executive Order 181 in New York State. Signed in April of 2018, the Executive Order restored voting rights to many individuals on parole. Prior to Executive Order 181, individuals were disenfranchised while incarcerated or on parole. Most individuals who were on parole on Election Day in 2018 were allowed to vote, a major step forward for democracy in the Empire State. In this paper, however, I focus on individuals who were discharged from parole prior to Election Day: individuals who would not have been directly disenfranchised by felony disenfranchisement policies even in the absense of the Executive Order. I hypothesize that Executive Order 181 increased turnout among this population, even though their eligibility was not directly impacted by the policy change.

@@

Secondly, Executive Order 181 is expected to dispel confusion about eligibilty. Since @Downs1957, political scientists have argued that an individual's propensity to vote is decided at least in part by comparing the expected costs of casting a ballot with the expected benefits of doing so. The costs of voting while disenfranchised (intentionally or otherwise) are exceedingly high, as high-profile cases in states such as Texas have made clear in recent years [@Flynn2018]. If an individual assessment of the cost of voting is determined in part by the product of their uncertainty about their eligibility and the cost of casting a ballot while ineligible, formerly disenfranchised individals are unlikely to participate unless they are fully certain of their eligibility. As @Drucker2005 and others have detailed, however, many formerly incarcerated individuals are misinformed about their eligibility to cast a ballot. By dispelling uncertainty around eligibility, I hypothesize that Executive Order 181 increased turnout among the formerly disenfranchised.

@@

On April 18^th^, 2018, Governor Andrew Cuomo signed Executive Order 181 which effectively ended the disenfranchisement of New Yorkers on parole. While the Executive Order is obviously beneficial for individuals who are on parole on Election Day (and therefore would have been disenfranchised but for the policy change), the Executive Order may have benefits even for the individuals whose eligibility was not directly impacted. Under the Executive Order, parole officers were required to inform their parolees of their new status under the law. In the analysis that follows, I look only at individuals who were discharged before October 10^th^, 2018 (the registration deadline in New York State). Limiting the analysis to individuals whose eligibility to cast a ballot was not directly impacted by the policy change allows me to explore whether having one's rights restored while still under formal supervision is effective at raising that individual's propensity to vote even once supervision is ended.

In order to estimate the impact of Executive Order 181 on turnout, I begin by exploring whether the turnout rate among individuals discharged from parole after the policy went into effect systematically differed from those discharged earlier. This analysis uses an individual-level logistic regression that estimates individuals' turnout rates and controls for various individual-level characteristics.

@@

## Voter File Data {-}
Most states in the United States are required to maintain files with information on all registered voters. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first, middle, and last name; date of birth; vote history; and other information. The New York State Voter File also includes information on voters who were previously registered but have since been purged, either because they moved, died, or were incarcerated for a felony offense. I use a snapshot of the registered voter file from March 3^rd^, 2019.

## Matching {-}
Turnout in the 2018 midterm election is estimated by matching the parole records with the registered voter file. I match individuals in each dataset using first name, middle name, last name, and date of birth. To be considered a "match," records must have the exact same birth date. The first and last names must also be exact matches (conditional on the adjustments discussed below). The middle names must meet one of the following conditions in order to qualify:

* Middle names are identical. If neither set of records includes a middle name, this condition is met.
* A full middle name in one set of records and only a middle initial in the other. The first letter of the full middle name must be the same as the middle initial in the other set of records.
* A middle name or middle initial in one set of records, and a missing middle name in the other set.

Thus, "John Andrew Doe" and "John A Doe" would count as matches. Similarly, "John Andrew Doe" and "John Doe" would count, while "John Andew Doe" and "John Anthony Doe" would not.

There are two types of potential error in this methdology: a false positive will result when a parolee's records matches the record of a voter who is a different individual but shares the same name and date of birth. False negatives will occur when an individual has a different name in the different sets of records, or when the birthdate is incorrectly reported in one of the sets of records

Testing for the presence of false positive matches is fairly straightforward. @Meredith2013 offers one way to test their prevalence using placebo matching. I slightly alter the date of birth reported in the parole discharge dataset to create false records. Comparing the number of matches between these "fake" discharge records and the voter file with the number of matches between the "true" records and the voter file provides an estimate of how frequently false positives occur. Table \ref{tab:change-dobs} shows the results of true matches, as well as when I construct a set of fake records by adding or subtracting 35 days from a parolee's birthdate. This analysis indicates that false positives account for between 0.6 and 0.7 percent of all matches, a share that is likely too small to have any material impact on the overall analysis.

```{r shift-dobs-chunk, echo = F}
less_35 <- scales::comma(readRDS("./temp/m35_2.rds"))
true <- scales::comma(nrow(readRDS("./temp/din_nys_parolees.rds")))
plus_35 <- scales::comma(readRDS("./temp/p35_2.rds"))

table <- data.frame(group = c("Actual Birthdate", "Birthdate + 35 Days", "Birthdate - 35 Days"), count = c(true, plus_35, less_35))

j <- table %>% 
  mutate_all(linebreak) %>% 
  knitr::kable(booktabs = T, caption = "(\\#tab:change-dobs) Results of Shifting Birthdates", linesep = "", escape = F,
               col.names = linebreak(c("Group", "Number of Matches Between\nDOCCS and Voter File Records")),
               align = "c") %>% 
  kable_styling(latex_options = c("HOLD_position"),
                position = "center")
j

```
Testing for false negatives is more challenging. If an individual marries and changes her name after being discharged from parole, for instance, I will not identify her using my matching methodology. Similarly, "John Doe" and "Jonathan Doe" would not result in a match. To reduce the likelihood of these false negatives I remove all punctuation from all names, and standardize capitalization. A record with a last name of "O'Donnell" in one dataset, therefore, would match a last name of "O DONNELL" in the other (provided the other criteria are satisfied). Such standardizations, however, will miss individuals who change their names entirely. For three reasons, however, this is not likely to present major challenges: firstly, women are far more likely to change their last names than men, and women make up barely 6 percent of individuals who have been discharged from felony parole. Secondly, because both parolee discharge and voter registration are legal records, individuals are likely to be recorded using their full names (that is to say, an individual is unlikely to be "John" in one set of records and "Jonathan" in the other). Finally, rates of false negatives are likely to be constant within the state during the study period, and there is no reason to believe that these false negatives would be associated with being discharged from parole after the Executive Order went into effect.

@@

Below, I employ a logistic regression to investigate the impact of Executive Order 181 on turnout among all parolees discharged after the policy went into effect. Because not all individuals had their rights restored, I follow this analysis with an instrumental-variables approach to determine the effect of rights restoration on individuals whose rights were in fact formally restored prior to discharge from parole.

@@

```{r to-16-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-16}Turnout in 2016 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to16_chart.rds")
print(j)
```
Figure \ref{fig:to-18} plots month of parole discharge and turnout in the 2018 midterm elections. Once again, a weighted quadratic curve is fitted with a 95 percent confidence band. This curve is fit on monthly data running from January, 2012 through April, 2018, and extended through October, 2018.
```{r to-18-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18}Turnout in 2018 Midterm Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart.rds")
print(j)
```
Figure \ref{fig:to-16} does not indicate that individuals who were discharged from parole shortly before the 2016 presidential election were more likely to cast a ballot than individuals discharged earlier in the year. Figure \ref{fig:to-18}, on the other hand, indicates that New Yorkers discharged from parole in the months leading up to the 2018 election --- many of whom had their rights restored while they were still on parole --- were more likely to participate than those discharged earlier in the year. However, Figures \ref{fig:to-16} and \ref{fig:to-18} are noisy and do not prove that Executive Order increased turnout.

@@

Although Figures \ref{fig:to-16} and \ref{fig:to-18} indicate that the length of time an individual has been off parole is correlated with her propensity to vote, that relationship is far weaker in the short term. Figure \ref{fig:to-18-recent} indicates that parole discharge date and turnout rates are not correlated when we limit the analysis to individuals discharged in 2017 or 2018.

```{r to-18-recent-chart, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:to-18-recent}Turnout in 2018 Presidential Election", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/to18_chart_recent.rds")
print(j)
```
Formalizing this chart into an individual-level logit model demonstrates that time since discharge is not correlated with turnout in the short run. The models in Table \ref{tab:to-18-logit-short} include individuals last discharged from parole between January 1^st^, 2017, and April 17^th^, 2018.

@@

The inclusion of time controls in Table \ref{tab:to-18-logit-short} increases the AIC. A Chi-squared test confirms that the model is not improved when controls for time are included.^[[Appendix A] provides further corroboration that being discharged from parole in the months before an election is uncorrelated with propensity to vote by exploring turnout rates in the 2016 presidential election.] Over the short term, the length of time an individual has been off parole is not associated with his propensity to vote.

This lack of correlation between discharge date and turnout indicates that we can conceptualize Executive Order as a natural experiment. We can expect that, in the absence of Executive Order 181, individuals discharged from parole before and after the policy went into effect would have turned out at comparable rates. Any observed difference in turnout rates, therefore, can be attributed to the implementation of the policy.

Although Governor Cuomo signed the Executive Order on April 18, 2018, an examination of the individuals whose rights were ultimately restored indicates that the program did not go into full effect until later in May. Figure \ref{fig:restoration-chart} plots the share of individuals discharged from parole each day in the spring of 2018. It is not until May 21^st^ that the majority of discharged parolees had their voting rights restored prior to their discharge.
```{r restoration-chart-chunk, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 4, fig.cap = "\\label{fig:restoration-chart}Share of Discharged Parolees Whose Voting Rights Were Restored Prior to Discharge", fig.pos = "H", fig.align = "center"}
j <- readRDS("./temp/restoration_plot.rds")
print(j)
```

In Table \ref{tab:to-18-logit}, I present the results of an individual-level logistic regression exploring turnout in 2018 among individuals who were discharged on or after May 2121^st^, 2018, the date when the policy went into full effect. It includes all individuals discharged from parole between January 1^st^, 2017, through October 10^th^, 2018 (the registration deadline in New York State).

```{r logit-reg-chunk, echo = F, results = "asis", message = F}
load("./temp/individual_turnout_18_itt.rdata")

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Discharged After EO 181)", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2018 Election"),
          title = "(\\#tab:to-18-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```

Model 1 in Table \ref{tab:to-18-logit} formalizes the trend presented in Figure \ref{fig:to-18-recent-chart} by controlling only for whether an individual was discharged after Executive Order 181 went into effect. Model 2 also controls for individual-level characteristics: sex, age on November 6^th^, 2018, county of incarceration, and race. Model 3 adds sentence-specific information to Model 2: the number of counts in the individual's most recent sentence, the amount of time they spent on parole, and the class(es) of felony for which they were convicted. In each successive model, the AIC decreases substantially, indicating that these controls are warranted. Table \ref{tab:to-18-logit} makes clear that formerly incarcerated men were far less likely to vote than formerly incarcerated women; that older formerly incarcerated individuals were more likely to cast a ballot; and individuals who spent longer on parole were more likely to participate in the midterm election.

Each model also indicates that individuals discharged from parole after the Executive Order went into effect were more likely to cast a ballot than those discharged earlier. Exponentiating the coefficients on D(Discharged After EO 181) indicates that Executive Order 181 raised the turnout rate among formerly disenfranchised individuals by between `r round((exp(model1[["coefficients"]][["finished_postTRUE"]]) - 1) * 100, digits = 1)` and `r round((exp(model3[["coefficients"]][["finished_postTRUE"]]) - 1) * 100, digits = 1)` percent.

Table \ref{tab:to-18-logit} indicates that Executive Order was successful at increasing turnout among all formerly disenfranchised individuals. The above analysis, however, likely understates the impact of the Executive Order on individuals who were in fact eligible to cast a ballot. As Figure \ref{fig:restoration-chart} indicates, not all individuals discharged after the Executive Order went into effect had their rights restored. Individuals who were discharged after the policy was changed but but did not have their rights restored can be conceptualized as "non-compliers" --- although they were "assigned" to the treatment group (were discharged after the policy changed), they did not receive the treatment (have their rights formally restored). Some of these non-compliers were likely non-citizens, who could not be granted voting rights. Others may have been re-arrested, which disqualified them from having rights restored. Both non-citizens and individuals who have been recently arrested are unlikely to vote. "Compliance," therefore, is almost certainly correlated with with an individual's likelihood of participating in the 2018 election. As such, the estimated treatment effect on the treated --- on individuals who were likely eligible to cast a ballot on election day --- requires an instrumental variables approach.

In the models below, a dummy indicating whether an individual was discharged from parole after the Executive Order went into effect (i.e., was in the "intend-to-treat" group) is used as an instrument for whether an individual's rights were actually restored (i.e., "complied" with the treatment). Table \ref{tab:iv-models} shows the second-stage estimation results, estimated using an OLS sepcification.^[Although the dependent variable remains dichotomous (as do both the independent variable of interest and the instrument), @Angrist2008 recommends using OLS in instrumental variables analyses even under such conditions.]

@@

Table \ref{tab:iv-models} indicates that using discharge date of on or after May 21^st^ as an instrument for rights restoration is warranted. The p-values of the Wu-Hausman tests hover just below 0.05, while the Weak Instruments test is highly significant in each model.

Model 3 in Table \ref{tab:iv-models} indicates that having one's rights restored caused an individual's propensity to vote to increase by `r round(summary(iv3)[["coefficients"]][2,1] * 100, digits = 2)` percentage points. Approximately `r round(readRDS("./temp/to_restored.rds") * 100, digits = 2)` percent of individuals who had their rights restored cast a ballot in 2018, indicating that Executive Order 181 increased the propensity to vote by around `r round(summary(iv3)[["coefficients"]][2,1] / (readRDS("./temp/to_restored.rds") - summary(iv3)[["coefficients"]][2,1]) * 100, digits = 1)` percent. Unsurprisingly, this estimate is higher than the estimates reported in Table \ref{tab:to-18-logit}.

## Discussion {-}
As discussed above, the effect of incarceration on individuals' propensity to vote is an open question for the literature. What is known, however, is that individuals who have been to prison were likely to vote at very low rates prior to incarceration, and continue to vote at very low rates after incarceration. Whether incarceration reduces political participation or not, increasing the propensity to vote of individuals who have been to prison is a laudable goal. Prisoners overwhelmingly come from marginalized communities with much to gain from policy. Executive Order 181 appears to have substantially boosted turnout among formerly incarcerated individuals through two mechanisms. The first is obvious: parolees who would not have been eligible to vote absent the Executive Order were allowed to cast ballots in the midterm elections. The second mechanism is less obvious. Individuals who finished parole between May 21^st^, 2018, and  October 10^th^, 2018, would have been eligible to participate in the election even if the governor had not signed the Executive Order. The EO 181, however, appears to have boosted turnout even among individuals whose eligibility to vote was not directly impacted by the rules change. The increase in turnout was substantial: Executive Order 181 increased turnout among individuals who had their rights restored by more than 35 percent, and raised the turnout of formerly disenfranchised individuals as a whole by

@@

# Appendix A {-}

In the [Individual-Level Turnout Regressions] section of this paper, I argue that being discharged from parole in the final months leading up to an election is uncorrelated with propensity to vote. Table \ref{tab:to-16-logit} demonstrates that individuals discharged between May 21^st^ -- October 14^th^, 2016, did not participate at different rates in the 2016 presidential election than other formerly incarcerated individuals. Table \ref{tab:to-16-logit} includes all individuals last discharged from parole between January 1^st^, 2015, and October 14^th^, 2016.
```{r logit-reg-chunk-16, echo = F, results = "asis", message = F}
load("./temp/individual_turnout_16.rdata")

stargazer(model1, model2, model3,
          header = F,
          type = "latex",
          covariate.labels = c("D(Discharged on or after May 21\\textsuperscript{st}, 2016)", "Days Since Discharged",
                               "Days Since Discharged\\textsuperscript{2}", "D(Male)", "Age (Years)", "Counts in Most Recent Sentence",
                               "Time Spent on Parole (Years)"),
          dep.var.labels = c("Cast Ballot in 2016 Election"),
          title = "(\\#tab:to-16-logit) Individual-Level Logit Model",
          table.placement = "H",
          omit = c("county", "race", "felony"),
          add.lines=list(c("County Fixed Effects", "", "X", "X"),
                         c("Race / Ethnicity Fixed Effects", "", "X", "X"),
                         c("Felony Class Fixed Effects", "", "", "X")),
          table.layout = "-dm#-t-a-s-n")
```
Although turnout was generally higher in 2016 than in 2018 (reflecting statewide higher turnout thanks to the presidential contest), there is no evidence that being discharged in the summer of 2016 was associated with an individual's propensity to cast a ballot (the p-value on the coefficient of interest exceeds 0.5 in each model). The nonsignificant results from 2016 provide strong corroboration that discharge date serves as an effective instrument for rights restoration.
